---
title: "SPRING: Semi-Parametric Rank-Based Microbial Network Estimation"
author:
  - name: Grace Yoon
  - name: Irina Gaynanova
  - name: Christian Mueller
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{SPRING: Semi-Parametric Rank-Based Microbial Network Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

SPRING (Semi-Parametric Rank-based approach for INference in Graphical models)
estimates sparse microbial association networks from microbiome count data. The
method is designed to handle the key challenges of microbiome data: excess zeros
and compositionality.

The main steps are:

1. Optionally normalize compositional data using the modified centered log-ratio
   (mclr) transformation.
2. Estimate a rank-based latent correlation matrix via a truncated Gaussian
   copula model using `latentcor`.
3. Fit a sparse graphical model using Meinshausen-Buehlmann (MB) neighborhood
   selection across a regularization path.
4. Select the optimal sparsity level via StARS (Stability Approach to
   Regularization Selection) subsampling.

For full methodological details see Yoon, Gaynanova and Mueller (2019)
*Frontiers in Genetics* 10:516.
[doi:10.3389/fgene.2019.00516](https://doi.org/10.3389/fgene.2019.00516)

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SPRING")
```

# Example data

The package includes `QMP`, a quantitative microbiome profiling dataset from
Vandeputte et al. (2017) with 106 subjects and 91 genera, used as the primary
benchmark in the original paper.

```{r data}
library(SPRING)
data(QMP)
dim(QMP)
```

# Running SPRING

## Quantitative count data

When absolute counts are available (`quantitative = TRUE`), no normalization is
applied. We use a short lambda sequence and few subsamples here for speed; the
paper uses `nlambda = 50` and `rep.num = 50`.

```{r spring-quantitative}
set.seed(10010)
fit <- SPRING(QMP, quantitative = TRUE, nlambda = 15, rep.num = 20,
              seed = 10010, verbose = FALSE)
```

## Compositional (relative abundance) data

When only relative abundances are available (`quantitative = FALSE`, the
default), SPRING first applies the mclr transformation internally.

```{r spring-compositional, eval=FALSE}
RMP <- QMP / rowSums(QMP)
fit_comp <- SPRING(RMP, quantitative = FALSE, nlambda = 15, rep.num = 20,
                   seed = 10010, verbose = FALSE)
```

## Data-specific lambda sequence

Setting `lambdaseq = "data-specific"` generates a lambda sequence scaled to the
estimated correlation structure of the data, which can improve edge recovery.

```{r spring-dataspecific, eval=FALSE}
fit_ds <- SPRING(QMP, quantitative = TRUE, lambdaseq = "data-specific",
                 nlambda = 15, rep.num = 20, seed = 10010, verbose = FALSE)
```

# Extracting the network

The optimal regularization index selected by StARS and the corresponding
adjacency matrix:

```{r network}
opt.index <- fit$output$stars$opt.index
cat("Optimal lambda index:", opt.index, "\n")
cat("Optimal lambda:", fit$lambdaseq[opt.index], "\n")

# Adjacency matrix at optimal lambda
adj <- as.matrix(fit$fit$refit$stars)
cat("Number of edges:", sum(adj) / 2, "\n")
```

# Session information

```{r session}
sessionInfo()
```
