---
title: "SPRING: Semi-Parametric Rank-Based Microbial Network Estimation"
author:
  - name: Grace Yoon
  - name: Irina Gaynanova
  - name: Christian Müller
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{SPRING: Semi-Parametric Rank-Based Microbial Network Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

SPRING (Semi-Parametric Rank-based approach for INference in Graphical models)
estimates sparse microbial association networks from microbiome count data. The
method is designed for quantitative microbiome data, but can also work with compositional data via modified central log-ratio (mclr) transformation. 
For full methodological details, see Yoon G., Gaynanova I. and Müller C.L. (2019) [Microbial Networks in
SPRING - Semi-parametric Rank-Based Correlation and Partial Correlation
Estimation for Quantitative Microbiome
Data](https://www.frontiersin.org/articles/10.3389/fgene.2019.00516/full).
*Frontiers in Genetics*, 10:516.

The main steps are:

1. [For compositional data] Normalize data using the 
   *mclr* transformation.
2. Estimate a rank-based latent correlation matrix via a truncated Gaussian copula model. This is based on R package `latentcor` ([Huang et al. (2021)](https://doi.org/10.21105/joss.03634)).
3. Fit a sparse graphical model using Meinshausen-Buehlmann (MB) neighborhood
   selection across a regularization path.
4. Select the optimal sparsity level via StARS (Stability Approach to
   Regularization Selection) subsampling.


# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SPRING")
```

# Example data

The package includes `QMP`, a quantitative microbiome profiling dataset from
[Vandeputte et al. (2017)](https://doi.org/10.1038/nature24460) with 106
subjects and 91 genera, used as the primary benchmark in the original paper.

```{r data}
library(SPRING)
data(QMP)
dim(QMP)
```

# Running SPRING

## Quantitative count data

When absolute counts are available (`quantitative = TRUE`), no normalization is
applied. We use a short lambda sequence and few subsamples here for speed; the
paper uses `nlambda = 50` and `rep.num = 50`.

```{r spring-quantitative}
set.seed(10010)
fit <- SPRING(QMP, quantitative = TRUE, nlambda = 15, rep.num = 20,
              seed = 10010, verbose = FALSE)
```

## Compositional (relative abundance) data

When only relative abundances are available (`quantitative = FALSE`, the
default), SPRING first applies the mclr transformation internally.

```{r spring-compositional, eval=FALSE}
RMP <- QMP / rowSums(QMP)
fit_comp <- SPRING(RMP, quantitative = FALSE, nlambda = 15, rep.num = 20,
                   seed = 10010, verbose = FALSE)
```

## Data-specific lambda sequence

Setting `lambdaseq = "data-specific"` generates a lambda sequence scaled to the
estimated correlation structure of the data, which can improve edge recovery.

```{r spring-dataspecific, eval=FALSE}
fit_ds <- SPRING(QMP, quantitative = TRUE, lambdaseq = "data-specific",
                 nlambda = 15, rep.num = 20, seed = 10010, verbose = FALSE)
```

# Extracting the network

The optimal regularization index selected by StARS and the corresponding
adjacency matrix:

```{r network}
opt.index <- fit$output$stars$opt.index
cat("Optimal lambda index:", opt.index, "\n")
cat("Optimal lambda:", fit$lambdaseq[opt.index], "\n")

# Adjacency matrix at optimal lambda
adj <- as.matrix(fit$fit$refit$stars)
cat("Number of edges:", sum(adj) / 2, "\n")
```

# Visualizing the network

The estimated Meinshausen-Buehlmann (MB) coefficients at the optimal lambda
capture both the presence and direction of associations. We symmetrize them
(averaging $\hat{\beta}_{ij}$ and $\hat{\beta}_{ji}$) and focus on the
sub-network of taxa that have at least one strong association $\geq |0.2|$,
following the criterion used in the original paper.

```{r visualization, fig.width=7, fig.height=7}
library(corrplot)

# Run SPRING with paper settings (nlambda = 50, rep.num = 50)
set.seed(10010)
fit_ann <- SPRING(QMP, quantitative = TRUE, nlambda = 50, rep.num = 50,
                  seed = 10010, verbose = FALSE)
opt_ann <- fit_ann$output$stars$opt.index

# Symmetrize MB coefficients at optimal lambda
beta_opt <- as.matrix(fit_ann$fit$est$beta[[opt_ann]])
rownames(beta_opt) <- colnames(beta_opt) <- colnames(QMP)
beta_sym <- (beta_opt + t(beta_opt)) / 2

# Subset to taxa with at least one strong association >= |0.2|
strong <- apply(abs(beta_sym), 1, max) >= 0.2
beta_sub <- beta_sym[strong, strong]
cat("Taxa in sub-network:", sum(strong), "\n")
cat("Strong edges:", sum(abs(beta_sub) >= 0.2) / 2, "\n")

corrplot(beta_sub, method = "color", type = "upper",
         tl.cex = 0.6, tl.col = "black",
         title = "Sub-network: taxa with at least one |association| >= 0.2",
         mar = c(0, 0, 2, 0))
```

Blue entries indicate positive associations; red entries indicate negative
associations. Taxa without genus-level classification are labeled by the best
available higher-level taxonomy (family, order, class, or phylum), with numeric
suffixes disambiguating multiple unclassified taxa within the same rank (e.g.,
`Ruminococcaceae` and `Ruminococcaceae_1`).

**Note:** The sub-network shown here may differ slightly from Figure 10 of
Yoon et al. (2019) because this package now uses `latentcor` for latent
correlation estimation rather than the original `mixedCCA` implementation,
which can produce minor differences in estimated associations.

# Session information

```{r session}
sessionInfo()
```
